Data Specialization for a Staged Lambda Calculus

### 2014-11-06 paper/splitting.tex
First cut at the start of Stage 2 splitting.

Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
as evaluating $e$ in \lang.

@@

\subsubsection{Splitting}

The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

% There is one final subtlety to splitting stage \bbone\ terms. In the example
% above, there are \emph{two} flavors of stage \bbone\ subcomputations:
% \texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
% a precomputation which is used by the computation of the \bbtwo-mask.

Formally, splitting a stage \bbone\ term, as implemented by the judgment $e
\splitonesym [c,l.r]$ (\ref{fig:termSplitOne}), yields a \emph{combined result}
$c$ which computes a pair of the stage \bbone\ result and the precomputation,
and a \emph{resumer} function $l.r$. The correctness of splitting requires that,
for $e\mathbin{\redonesym}[\xi;v]$, we have that $\pio~c$ agrees with
$\masko{v}$, and $\letin l{\pit~c}r$ agrees with $\maskt{\xi;v}$.

The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
\texttt{((),12)}, which agrees with the stage \bbtwo\ mask.

\paragraph{Stage \bbone\ masking}

The purpose of masking is to expose only the relevant structure of a partial
value $v$ at each stage. For the most part, stage \bbone\ masking, $\masko{v}$,
simply replaces all $\next$s in $v$ with \texttt{()}. The only exception is for
functions $\lam xAe$; such a value represents a computation, so masking splits
its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
stage \bbone\ component, the combined result.

\paragraph{Stage \bbtwo\ masking}

Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
residual table $\xi$, because the stage \bbtwo\ parts of $v$ contain references
to $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
structure, and then reify the resulting term in the context $\xi$. This
rewriting process replaces base type constants in $v$ with \texttt{()}, and
$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

\paragraph{Splitting}

Splitting stage \bbone\ terms, as implemented by the judgment $e\splitonesym
[c,l.r]$, is also guided by the structure of the term $e$.

\TODO polish the following explanation

>>>>

Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)

@@

\subsubsection{Splitting Mechanics}

Given the definition of masking, splitting a stage \bbone\ term $\colone e A$ 
(where $e\mathbin{\redonesym}[\xi;v]$), yields a \emph{combined term}
$c$ that performs all stage~\bbone\ computations in $e$ and a \emph{resumer} function $l.r$ that represents all stage~\bbtwo\ computations.
Whereas in stage~\bbtwo\ term splitting,
$p$ constitutes only the precomputation to the resumer, in stage~\bbone\ splitting we require $c$ to reduce (via~$\redsym$) to the tuple $(y,z)$, where $y$ is an \emph{immediate result} that is identical to $\masko{\xi;v}$, and $z$ is the required precomputation such that $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.

Splitting stage~\bbone\ terms is implemented by the judgment $e \splitonesym [c,l.r]$ in \ref{fig:termSplitOne}.

% The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
% resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
% which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
% \texttt{((),12)}, which agrees with the stage \bbtwo\ mask.

\TODO REDO THE REST OF THE TEXT TO WALK THROUGH THE MAJOR RULES OF SPLITTING
